---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# frictionless

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/frictionless)](https://CRAN.R-project.org/package=frictionless)
[![R-CMD-check](https://github.com/frictionlessdata/frictionless-r/workflows/R-CMD-check/badge.svg)](https://github.com/frictionlessdata/frictionless-r/actions)
[![codecov](https://codecov.io/gh/frictionlessdata/frictionless-r/branch/main/graph/badge.svg?token=bKtiHW21K0)](https://codecov.io/gh/frictionlessdata/frictionless-r)
[![repo status](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![Funding](https://img.shields.io/static/v1?label=powered+by&message=lifewatch.be&labelColor=1a4e8a&color=f15922)](https://lifewatch.be/)
<!-- badges: end -->

Frictionless is an R package to read and write [Frictionless Data Packages](https://specs.frictionlessdata.io/data-package/) in a [Tidyverse](https://www.tidyverse.org/) way.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("frictionlessdata/frictionless-r")
```

## Usage

### Read data from a local Data Package

You can read data from a Data Package using the two read functions 
(`read_package()`, then `read_resource()`):

```{r example_local, message=FALSE}
library(frictionless)

# Read a datapackage.json file
package <- read_package(system.file("extdata", "datapackage.json", package = "frictionless"))

# List the resource names
package$resource_names

# Read data from the resource "observations"
read_resource(package, "observations")
```

### Read data from an online Data Package

The read functions can also be used to access all data from an online Frictionless Data Package (e.g. one published on [Zenodo](https://doi.org/10.5281/zenodo.5070086)) via the `datapackage.json` URL:

```{r example_zenodo}
# Read the datapackage.json file: this will give you access to all Data 
# Resources without reading them, which is convenient and fast
package <- read_package("https://zenodo.org/record/5070086/files/datapackage.json")

# List the resource names
package$resource_names

# Read the gps data: will return a single data frame, even though the data are 
# split over multiple csv files
read_resource(package, "gps")
```

For more functionality, see the [function reference](https://frictionlessdata.github.io/frictionless-r/reference/index.html).
