---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# datapackage

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/datapackage)](https://CRAN.R-project.org/package=datapackage)
[![R-CMD-check](https://github.com/inbo/datapackage/workflows/R-CMD-check/badge.svg)](https://github.com/inbo/datapackage/actions)
[![codecov](https://codecov.io/gh/inbo/datapackage/branch/main/graph/badge.svg?token=bKtiHW21K0)](https://codecov.io/gh/inbo/datapackage)
<!-- badges: end -->

datapackage is an R package to read and write [Frictionless Data Packages](https://specs.frictionlessdata.io/data-package/) in a [Tidyverse](https://www.tidyverse.org/) way.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("inbo/datapackage")
```

## Usage

There are two functions: `read_package()` and `read_resource()`. Basic functionality:

```{r example_local}
library(datapackage)

# Read datapackage.json file
package <- read_package(system.file("extdata", "datapackage.json", package = "datapackage"))

# List resource names
package$resource_names

# Read data from resource "observations"
read_resource(package, "observations")
```

Datapackage allows you to access all data from an external Frictionless Data Package (e.g. one published on [Zenodo](https://zenodo.org/search?page=1&size=20&q=frictionlessdata&type=dataset)) via the `datapackage.json` URL:

```{r example_zenodo}
# Read datapackage.json file: this will give you access to all data resources,
# without reading them, which is convenient and fast
package <- read_package("https://zenodo.org/record/5070086/files/datapackage.json")

# List resource names
package$resource_names

# Read gps data: will return a single Data Frame, even though data are 
# split over multiple csv files
read_resource(package, "gps")
```
