---
title: "Table Dialect support"
author: "Peter Desmet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Table Dialect support}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

[Table Dialect](https://specs.frictionlessdata.io/csv-dialect/) (previously called CSV dialect) is a simple format to describe the dialect of a tabular data file, including its delimiter, header rows, escape characters, etc.

_In this document we use the terms "package" for Data Package, "resource" for Data Resource, "dialect" for Table Dialect, and "schema" for Table Schema._

## Support

Frictionless supports most dialect properties to read [Tabular Data Resources](https://specs.frictionlessdata.io/tabular-data-resource/). Dialect manipulation is limited to setting a `delimiter`. When writing resources, it (mainly) makes uses of default dialect properties, removing the necessity to define them.

### Read

`read_resource()` uses the `dialect` property of a resource to parse a tabular data file. Only properties that deviate from the default need to be specified. E.g. a tab-delimited file without header rows must have the following dialect:

```json
"dialect": {
  "delimiter": "\t",
  "header": false
}
```

### Manipulate

Frictionless does not support direct manipulation of the dialect. `add_resource()` allows to set one property (`dialect$delimiter`) when data are provided as a file, all other properties are assumed to be the default.

### Write

`write_package()` writes a package to disk as a `datapackage.json` file. This file will include the metadata of all the resources, including the dialect (if defined). `write_package()` writes resources created from a data frame to CSV files, but no `dialect` property is set for those, since only defaults are used.

## Dialect properties

### delimiter

[`delimiter`](https://specs.frictionlessdata.io/csv-dialect/#specification) is used by `read_resource()` and defaults to `","`. It is passed to `delim` in `readr::read_delim()`. `add_resource()` does not set `delimiter`, unless provided in `delim` and different from the default `","`:

```{r}
library(frictionless)
package <- example_package()

path <- system.file("extdata", "deployments.csv", package = "frictionless")
package <- add_resource(package, "deployments", data = path, delim = ",", replace = TRUE)
package$resources[[1]]$dialect$delimiter
# TODO: update to observations_1.tsv
```

### lineTerminator

[`lineTerminator`](https://specs.frictionlessdata.io/csv-dialect/#specification) is ignored by `read_resource()`. It relies on `readr::read_delim()` instead, which interprets line terminator `LF` and `CRLF` automatically and does not support `CR` (used by Classic Mac OS, final release 2001).

### quoteChar

[`quoteChar`](https://specs.frictionlessdata.io/csv-dialect/#specification) is used by `read_resource()` and defaults to `"`. It is passed to `quote` in `readr::read_delim()`.

### doubleQuote

[`doubleQuote`](https://specs.frictionlessdata.io/csv-dialect/#specification) is used by `read_resource()` and defaults to `true`, but can be overruled by `escapeChar`. It is passed to `escape_double` in `readr::read_delim()`.

### escapeChar

[`escapeChar`](https://specs.frictionlessdata.io/csv-dialect/#specification) is ignored by `read_resource()` unless it is `"\\"`. It is passed as `escape_backslash = TRUE` and `escape_double = FALSE` in `readr::read_delim()`. Note that `escapeChar` and `doubleQuote` are mutually exclusive, so you cannot escape with `\"` and `""` in the same file.

### nullSequence

[`nullSequence`](https://specs.frictionlessdata.io/csv-dialect/#specification) is ignored by `read_resource()`. Provide as `missingValues` in the schema instead (see `vignette("table-schema")`).

### skipInitialSpace

[`skipInitialSpace`](https://specs.frictionlessdata.io/csv-dialect/#specification) is used by `read_resource()` and defaults to `false`. It is passed to `trim_ws` in `readr::read_delim()`.

### header

[`header`](https://specs.frictionlessdata.io/csv-dialect/#specification) is used by `read_resource()` and defaults to `true`. It is passed as `trim_ws = 1` (or `0`) in `readr::read_delim()`.

### commentChar

[`commentChar`](https://specs.frictionlessdata.io/csv-dialect/#specification) is used by `read_resource()` and defaults to undefined. It is passed to `comment` in `readr::read_delim()`.

### caseSensitiveHeader

[`caseSensitiveHeader`](https://specs.frictionlessdata.io/csv-dialect/#specification) is ignored by `read_resource()`.

### csvddfVersion

[`csvddfVersion`](https://specs.frictionlessdata.io/csv-dialect/#specification) is ignored by `read_resource()`.
