---
title: "Table Dialect"
author: "Peter Desmet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Table Dialect}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

[Table Dialect](https://datapackage.org/standard/table-dialect/) is a simple format to describe the dialect of a tabular data file, including its delimiter, header rows, escape characters, etc.

::: {.callout-note}
In this document we use the terms "package" for Data Package, "resource" for Data Resource, "dialect" for Table Dialect, and "schema" for Table Schema.
:::

## General implementation

frictionless supports most dialect properties to read [Tabular Data Resources](https://datapackage.org/standard/data-resource/#tabular), but only those designed for [delimited formats](https://datapackage.org/standard/table-dialect/#delimited) (not structured, spreadsheet, database formats). Dialect manipulation is limited to setting a `delimiter`. When writing resources, it (mainly) makes uses of default dialect properties, removing the necessity to define them.

### Read

`read_resource()` uses the `dialect` property of a resource to parse a tabular data file. Only properties that deviate from the default need to be specified. E.g. a tab-delimited file without header rows must have the following dialect:

```json
"dialect": {
  "delimiter": "\t",
  "header": false
}
```

### Manipulate

frictionless does not support direct manipulation of the dialect. `add_resource()` allows to set one property (`delimiter`) when data are provided as a file, all other properties are assumed to be the default.

### Write

`write_package()` writes a package to disk as a `datapackage.json` file. This file includes the metadata of all the resources, including the dialect (if defined). `write_package()` writes resources created from a data frame to CSV files, but no `dialect` property is set for those, since only defaults are used.

## Properties implementation

### $schema

[`$schema`](https://datapackage.org/standard/table-dialect/#dollar-schema) indicates what version of the Table Dialect standard is used (v1 if undefined). **All dialect properties supported by frictionless are the same in v1 and v2**. `$schema` is therefore ignored by `read_resource()`. While `add_resource()` creates a v2 resource, it typically does not define a dialect, except to declare a non-default `delimiter` (same in v1 and v2). It therefore does _not_ set `$schema` to `"https://datapackage.org/profiles/2.0/tabledialect.json"` (the recommended v2 value). Neither does `upgrade_package()` which leaves the dialect as is for all resources.

### header

[`header`](https://datapackage.org/standard/table-dialect/#header) is used by `read_resource()` and defaults to `true`. It is passed as `skip = 1` (or `0` for `false`) in `readr::read_delim()`, ignoring the header. Field names in `schema` are used instead.

### headerRows


[`headerRows`](https://datapackage.org/standard/table-dialect/#headerRows) is ignored by `read_resource()`.

### headerJoin

<!-- Not spec compliant -->

[`headerJoin`](https://datapackage.org/standard/table-dialect/#headerJoin) is ignored by `read_resource()`.

### commentRows

<!-- Not spec compliant -->

[`commentRows`](https://datapackage.org/standard/table-dialect/#headerJoin) is ignored by `read_resource()`.

### commentChar

[`commentChar`](https://datapackage.org/standard/table-dialect/#commentChar) is used by `read_resource()` and defaults to undefined. It is passed to `comment` in `readr::read_delim()`.

### delimiter

[`delimiter`](https://datapackage.org/standard/table-dialect/#delimiter) is used by `read_resource()` and defaults to `","`. It is passed to `delim` in `readr::read_delim()`. `add_resource()` does not set `delimiter`, unless provided in `delim` and different from the default `","`:

```{r}
library(frictionless)
package <- example_package()

path <- system.file("extdata", "v2", "observations_1.tsv", package = "frictionless")
package <- add_resource(package, "observations", data = path, delim = "\t", replace = TRUE)
resource(package, "observations")$dialect$delimiter
```

### lineTerminator

[`lineTerminator`](https://datapackage.org/standard/table-dialect/#lineTerminator) is ignored by `read_resource()`. It relies on `readr::read_delim()` instead, which interprets line terminator `LF` and `CRLF` automatically and does not support `CR` (used by Classic Mac OS, final release 2001).

### quoteChar

[`quoteChar`](https://datapackage.org/standard/table-dialect/#quoteChar) is used by `read_resource()` and defaults to `"`. It is passed to `quote` in `readr::read_delim()`.

### doubleQuote

[`doubleQuote`](https://datapackage.org/standard/table-dialect/#doubleQuote) is used by `read_resource()` and defaults to `true`, but can be overruled by `escapeChar`. It is passed to `escape_double` in `readr::read_delim()`.

### escapeChar

[`escapeChar`](https://datapackage.org/standard/table-dialect/#escapeChar) is ignored by `read_resource()` unless it is `"\\"`. It is passed as `escape_backslash = TRUE` and `escape_double = FALSE` in `readr::read_delim()`.

::: {.callout-warning}
`escapeChar` and `doubleQuote` are mutually exclusive, so you cannot escape with `\"` and `""` in the same file.
:::

### nullSequence

[`nullSequence`](https://datapackage.org/standard/table-dialect/#nullSequence) is ignored by `read_resource()`. Provide as `missingValues` in the schema instead (see `vignette("table-schema")`).

### skipInitialSpace

[`skipInitialSpace`](https://datapackage.org/standard/table-dialect/#skipInitialSpace) is used by `read_resource()` and defaults to `false`. It is passed to `trim_ws` in `readr::read_delim()`.
