---
title: "Table Schema"
author: "Peter Desmet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Table Schema}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

[Table Schema](https://specs.frictionlessdata.io/table-schema/) is a simple format to describe tabular data, including field names, types, constraints, missing values, foreign keys, etc.

::: {.callout-info}
In this document we use the terms "package" for Data Package, "resource" for Data Resource, "dialect" for Table Dialect, and "schema" for Table Schema.
:::

## General implementation

Frictionless supports `schema$fields` and `schema$missingValues` to parse data types and missing values when reading [Tabular Data Resources](https://specs.frictionlessdata.io/tabular-data-resource/). Schema manipulation is limited to extracting a schema from a resource, creating one from a data frame, and providing one back to a resource. Schema metadata is including when writing a package.

### Read

`schema()` extracts the schema from a resource:

```{r}
library(frictionless)
package <- example_package()

# Get the Table Schema for the resource "observations"
schema <- schema(package, "observations")
str(schema)
```

`read_resource()` uses `schema$fields` to parse the names and data types of the columns in a tabular data file. For example, the third field in the schema (`timestamp`) is defined as a datetime `type` with a specific `format`:

```{r}
str(schema$fields[[3]])
```

`read_resource()` uses that information to correctly parse the data type and to assign the name `timestamp` to the column:

```{r}
observations <- read_resource(package, "observations")
observations$timestamp
```

The sixth field `life_stage` has an `enum` defined as one of its `constraints`:

```{r}
str(schema$fields[[6]])
```

`read_resource()` uses that information to parse the column as a factor, using `enum` as the factor levels:

```{r}
class(observations$life_stage)
levels(observations$life_stage)
```

### Manipulate

A schema is a list which you can manipulate, but frictionless does not provide functions to do that. Use `{purrr}` or base R instead (see `vignette("frictionless")`). You do not have to start a schema from scratch though: use `schema()` (see above) or `create_schema()` instead.

`create_schema()` creates a schema from a data frame and defines the `name`, `type` (and if a factor `constraints$enum`) for each field:

```{r}
# Create a schema from the built-in dataset "iris"
iris_schema <- create_schema(iris)
str(iris_schema)
```

`add_resource()` allows to include the schema with a resource. If no schema is provided, one is created with `create_schema()`:

```{r}
package <- add_resource(
  package,
  resource_name = "iris",
  data = iris,
  schema = iris_schema
)
```

### Write

`write_package()` writes a package to disk as a `datapackage.json` file. This file includes the metadata of all the resources, including the schema. To directly write a schema to disk, use `jsonlite::write_json()`.

## Schema properties implementation

### fields

[`fields`](https://specs.frictionlessdata.io/table-schema/#field-descriptors) is required. It is used by `read_resource()` to parse the names and data types of the columns in a tabular data file. `create_schema()` sets `fields` based on information in a data frame. See [Field properties implementation](#field-properties-implementation) for details.

<!-- ### $schema -->

<!-- ### fieldMatch -->

### missingValues

[`missingValues`](https://specs.frictionlessdata.io/table-schema/#missing-values) is used by `read_resource()` and defaults to `""`. It is passed to `na` in `readr::read_delim()`. `create_schema()` does not set `missingValues`. `write_package()` converts `NA` values to `""` when writing a data frame to a CSV file. Since this is the default, no `missingValues` property is set.

### primaryKey

[`primaryKey`](https://specs.frictionlessdata.io/table-schema/#primary-key) is ignored by `read_resource()` and not set by `create_schema()`.

<!-- ### uniqueKeys -->

### foreignKeys

[`foreignKeys`](https://specs.frictionlessdata.io/table-schema/#foreign-keys) is ignored by `read_resource()` and not set by `create_schema()`.

## Field properties implementation

### name

[`name`](https://specs.frictionlessdata.io/table-schema/#name) is used by `read_resource()` to assign a column name. The vector of names is passed as `col_names` to `readr::read_delim()`, ignoring names provided in the header of the data file. `create_schema()` uses the data frame column name to set `name`.

### type and format

[`type`](https://specs.frictionlessdata.io/table-schema/#types-and-formats) and (for some types) [`format`](https://specs.frictionlessdata.io/table-schema/#types-and-formats) is used by `read_resource()` to understand the column type. The vector of types is passed as `col_types` to `readr::read_delim()`, which warns if there are parsing issues (inspect with `problems()`). `create_schema()` uses the data frame column type to set `type`. See [Field types implementation](#field-types-implementation) for details.

`read_resource()` interprets `type` as follows:

field type | column type
--- | ---
[`string`](#string) | `character` or `factor`
[`number`](#number) | `double` or `factor`
[`integer`](#integer) | `double` or `factor`
[`boolean`](#boolean) | `logical`
[`object`](#object) | `character`
[`array`](#array) | `character`
[`datetime`](#datetime) | `POSIXct`
[`date`](#date) | `Date`
[`time`](#time) | `hms::hms()`
[`year`](#year) | `Date`
[`yearmonth`](#yearmonth) | `Date`
[`duration`](#duration) | `character`
[`geopoint`](#geopoint) | `character` 
[`geojson`](#geojson) | `character`
[`any`](#any) | `character`
other value | error
undefined | guessed

`create_schema()` sets `type` as follows:

column type | field type
--- | ---
`character` | `string`
`Date` | `date`
`difftime` | `number`
`factor` | `string` with factor levels as `constraints$enum`
`hms::hms()` | `time`
`integer` | `integer`
`logical` | `boolean`
`numeric` | `number`
`POSIXct`/`POSIXlt` | `datetime`
any other type | `any`

`create_schema()` does not set a `format`, since defaults are used for all types. This is also the case for datetimes, dates and times, since `readr::write_csv()` used by `write_package()` formats those to ISO8601, which is considered the default.

### title

[`title`](https://specs.frictionlessdata.io/table-schema/#title) is ignored by `read_resource()` and not set by `create_schema()`.

### description

[`description`](https://specs.frictionlessdata.io/table-schema/#description) is ignored by `read_resource()` and not set by `create_schema()`.

### example

[`example`](https://specs.frictionlessdata.io/table-schema/#example) is ignored by `read_resource()` and not set by `create_schema()`.

### constraints

[`constraints`](https://specs.frictionlessdata.io/table-schema/#constraints) is ignored by `read_resource()` and not set by `create_schema()`, except for `constraints$enum`. `read_resource()` uses it set the column type to `factor`, with `enum` values as factor levels. `create_schema()` does the reverse.

<!-- ### categories -->

<!-- ### categoriesOrdered -->

<!-- ### missingValues -->

### rdfType

[`rdfType`](https://specs.frictionlessdata.io/table-schema/#rich-types) is ignored by `read_resource()` and not set by `create_schema()`.

## Field types implementation

### string

[`string`](https://specs.frictionlessdata.io/table-schema/#string) is interpreted as `character`. Or `factor` when `constraints$enum` is defined.

- `format` is ignored.

### number

[`number`](https://specs.frictionlessdata.io/table-schema/#number) is interpreted as `double`. Or `factor` when `constraints$enum` is defined.

- `bareNumber` is supported. If `false`, whitespace and non-numeric characters are ignored.
- `decimalChar` (`.` by default) is supported, but as a single value for all number fields. If different values are defined, the most occurring one is selected.
- `groupChar` (undefined by default) is supported, but as a single value for all number fields. If different values are defined, the most occurring one is selected.

### integer

[`integer`](https://specs.frictionlessdata.io/table-schema/#integer) is interpreted as `double` (to avoid issues with big numbers). Or `factor` when `constraints$enum` is defined.

- `bareNumber` is supported. If `false`, whitespace and non-numeric characters are ignored.

### boolean

[`boolean`](https://specs.frictionlessdata.io/table-schema/#boolean) is interpreted as `logical`.

- `trueValues` that are not defaults are not supported.
- `falseValues` that are not defaults are not supported.

### object

[`object`](https://specs.frictionlessdata.io/table-schema/#object) is interpreted as `character`

### array

[`array`](https://specs.frictionlessdata.io/table-schema/#array) is interpreted as `character`.

<!-- ### list -->

### datetime

[`datetime`](https://specs.frictionlessdata.io/table-schema/#datetime) is interpreted as `POSIXct`.

- `format` is supported for the values `default` (ISO datetime), `any` (ISO datetime) and the same patterns as for `date` and `time`. The value `%c` is not supported.

### date

[`date`](https://specs.frictionlessdata.io/table-schema/#date) is interpreted as `Date`.

- `format` is supported for the values `default` (ISO date), `any` (guess `ymd`) and [Python/C strptime](https://docs.python.org/2/library/datetime.html#strftime-strptime-behavior) patterns, such as `%a, %d %B %Y` for `Sat, 23 November 2013`. `%x` is interpreted as `%m/%d/%y`. The values `%j`, `%U`, `%w` and `%W` are not supported.

### time

[`time`](https://specs.frictionlessdata.io/table-schema/#time) is interpreted as `hms::hms()`.

- `format` is supported for the values `default` (ISO time), `any` (guess `hms`) and [Python/C strptime](https://docs.python.org/2/library/datetime.html#strftime-strptime-behavior) patterns, such as `%I%p%M:%S.%f%z` for `8AM30:00.300+0200`.

### year

[`year`](https://specs.frictionlessdata.io/table-schema/#year) is interpreted as `Date` with month and day set to `01`.

### yearmonth

[`yearmonth`](https://specs.frictionlessdata.io/table-schema/#yearmonth) is interpreted as `Date` with day set to `01`.

### duration

[`duration`](https://specs.frictionlessdata.io/table-schema/#duration) is interpreted as `character`. You can parse these values with `lubridate::duration()`.

### geopoint

[`geopoint`](https://specs.frictionlessdata.io/table-schema/#geopoint) is interpreted as `character`.

### geojson

[`geojson`](https://specs.frictionlessdata.io/table-schema/#geojson) is interpreted as `character`.

### any

[`any`](https://specs.frictionlessdata.io/table-schema/#any) is interpreted as `character`
